### 2017-05-04 2017mergingmerges/section/05-Evaluation.tex
attempt to fit in sigconf 12p. format
serious cutting

\eject{}
% -----------------------------------------------------------------------------
\section{Fusion in Practice}
\label{s:Evaluation}

Stream fusion is ultimately performed for practical reasons. We want the fused result program to run faster than unfused source program. Fused result programs like the one in Fig. \ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell (maybe using Template Haskell), C or LLVM abstract assembly. The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.

% -----------------------------------------------------------------------------
\subsection{Fusibility}
\label{s:FusionOrder}
When we fuse a pair of processes we commit to a particular interleaving of instructions from each process. When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process. Consider our @alternatives@ example from \S\ref{s:EvaluationOrder}, here it is again:
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails. This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@. As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.

Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock. Statically, when we try to fuse the result process with the downstream @zip@ process then the deadlock is discovered and fusion fails. Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.\ref{fig:Fusion:Def:StepPair}. The $tryStepPair$ function determines which of the instructions from either process can be executed next, and when execution is deadlocked there are none.

@@

% -----------------------------------------------------------------------------
\subsection{Result Size}

As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together. The left of Fig.~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner. To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes. The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start. The right of Fig.~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream. In both cases the number of states in the result process grows linearly with the number of processes. In all combinations, with up to 7 processes there are less than 100 states in the result process. 
%%% AR: Note that the split version also has only one merge?

The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes. This is common with other systems based on inlining and/or template meta-programming, and is not prohibitive.

On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes. The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream. The output of each merge process is the input of the next, forming a chain of merges. In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.

When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product. During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition. The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result.

In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each. If we assume the two processes have about the same number of states then this is O($n^2$). Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused.

In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction. During fusion, the states for the consuming process are inlined at each occurrence of @push@. These states are legitimately different because at each occurence of @push@ the input channels of the merge process are in different channel states, and these channel states are included in the overall process state.

% -----------------------------------------------------------------------------
\subsection{Optimisation and Drop Instructions}
\label{s:Optimisation}
After we have fused two processes together, it may be possible to simplify the result before fusing in a third. Consider the result of fusing @group@ and @merge@ which we saw back in Fig.\ref{fig:Process:Fused}. At labels @F1@ and @F2@ are @jump@ instructions, where the first passes control to the second. The update expressions attached to these instructions are also non-interferring, which means we can safely combine these consecutive @jump@ instructions into a single @jump@ to @F4@. In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code. The (PreferJump) clauses of Fig.\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.

There are also @jump@ instructions like the one at @F5@ which have no associated update expressions, and thus can be eliminated completely. Another simple optimization is to perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.

Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has lower number of states. Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusibility of the overall process network will not be affected.

Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable. In Fig.\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction. Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable. Values pulled from stream @sIn1@ are first stored in @b1@ before being copied to @v@ and @x1@. When the two processes to be fused share a common input stream, use of stream buffer variable allows one process to continue using the value that was last pulled from the stream, while the other moves onto the next one.

When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable. This is the case in Fig.\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@. In the result, the interleaved instructions from both source processes then share the same heap variable.

@@

% -----------------------------------------------------------------------------
\begin{figure}
\begin{minipage}[t]{0.10\textwidth}
\end{minipage}
\begin{minipage}[t]{0.4\textwidth}
\begin{tikzpicture}
\begin{axis}[
% Hide the label on the second graph
	ylabel=Process states,
	xlabel=Pipelined combinators,
  ymin=0, ymax=100,
	enlargelimits=0.01,
	ybar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
\addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
\legend{With merge, Without merge};
\end{axis}
\end{tikzpicture}
\end{minipage}
\quad
\quad
\quad
\begin{minipage}[t]{0.4\textwidth}
\begin{tikzpicture}
\begin{axis}[
%       ylabel=Process states,
        xlabel=Parallel combinators,
  ymin=0, ymax=100,
        enlargelimits=0.01,
        ybar interval=0.7,
  width=7.5cm, height=4.5cm,
        legend pos=north west,
]
\addplot coordinates {(1,9) (2,27) (3,40) (4,53) (5,66) (6,79) (7,92)
  % Last bar doesn't show for some reason, so need to add a dummy value for the next one
    (8,1) };

\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };

\legend{With merge, Without merge};
\end{axis}
\end{tikzpicture}
\end{minipage}

\caption{Maximum output process size for fusing all combinations of up to $n$ combinators.}
\label{fig:bench:outputsize}
\end{figure}

% -----------------------------------------------------------------------------
\begin{figure}

\begin{tikzpicture}
\begin{axis}[
	ylabel=Process states,
	xlabel=Number of merges,
%  ymode=log,
  ymin=0, ymax=1500,
	enlargelimits=0.01,
	ybar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
% These are the values for splitting.
% They are smaller than the 'chaining', but look much nicer on the linear graph.
\addplot coordinates {(1,9) (2,42) (3,97) (4,196) (5,383) (6,746) (7,1461)
  % (8,2880)
  (8,1)
  };

% These are the values for chaining
% \addplot coordinates {(1,4) (2,48) (3,194) (4,760) (5,2814) (6,10064) (7,1) };

\end{axis}
\end{tikzpicture}

\caption{Exponential blowup occurs when splitting or chaining merges together.}
\label{fig:bench:exponential}
\end{figure}

>>>>

% -----------------------------------------------------------------------------
\section{Results}
\label{s:Evaluation}

Stream fusion is ultimately performed for practical reasons. We want the fused result program to run faster than unfused source program.

\subsection{Benchmarks}
We have implemented this system using Template Haskell in a library called @folderol@\footnote{\url{https://github.com/amosr/folderol}}.
To show practical examples, we use the finite stream extension mentioned in~\S\ref{s:Finite}.
We present three benchmarks: one array algorithm, and two simple file operations.

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
  symbolic y coords={Store, Recompute, Ours, Hand, end},
	xlabel=Runtime (ms),
  xmin=0, xmax=200,
	enlargelimits=0.01,
	xbar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
\addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
\end{axis}
\end{tikzpicture}
\caption{Runtime for Quickhull; lower is faster.}
\label{fig:bench:quickhull}
\end{figure}

Quickhull is a divide-and-conquer spatial algorithm to find the smallest convex hull containing all points.
At its core is an operation called `filterMax' which takes a line and an array of points, and finds the farthest point above the line, as well as all points above the line.

We compare hand-fused against ours, as well as two @Data.Vector@ versions, which uses shortcut fusion.
The shortcut fusion system cannot fuse both operations into a single loop, and both operations require the distances between the line and each point.
This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
We compared both, and recomputing was significantly faster.
However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.

Fig.~\ref{fig:bench:quickhull} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
The hand-fused version is the fastest, while our version is slower, but still faster than the two vector versions.
The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
  symbolic y coords={Conduit, Pipes, Streaming, Ours, end},
	xlabel=Runtime (ms),
  xmin=0, xmax=290,
	enlargelimits=0.01,
	xbar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
\addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
\end{axis}
\end{tikzpicture}
\caption{Runtime for append2; lower is faster.}
\label{fig:bench:append}
\end{figure}

The first file benchmark simply appends two files, while counting the number of lines.
We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
Fig.~\ref{fig:bench:append} shows the runtimes for appending 2MB of data.
The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
  symbolic y coords={Streaming, Pipes, Conduit hand, Ours, end},
	xlabel=Runtime (ms),
  xmin=0, xmax=1200,
	enlargelimits=0.01,
	xbar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
\addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };
\end{axis}
\end{tikzpicture}
\caption{Runtime for partition; lower is faster.}
\label{fig:bench:part}
\end{figure}

The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
The output lines are also counted.
The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or partially hand-fused.
Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
Fig.~\ref{fig:bench:part} shows the runtimes for partitioning a 6MB file.

% Fused result programs like the one in Fig. \ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell, C or LLVM abstract assembly. The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.

% -----------------------------------------------------------------------------
\subsection{Fusibility}
\label{s:FusionOrder}
When we fuse a pair of processes we commit to a particular interleaving of instructions from each process. When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process. Consider our @alternatives@ example from \S\ref{s:EvaluationOrder} again:

\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, this final fusion transform fails. This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@.
% As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.

Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock. Statically, when we try to fuse the result process with the downstream @zip@ process then the deadlock is discovered and fusion fails. Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.~\ref{fig:Fusion:Def:StepPair}. The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none.

@@

% -----------------------------------------------------------------------------
%\subsection{Result Size}
%
%As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together. The left of Fig.~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner. To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes. The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start. The right of Fig.~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream. In both cases the number of states in the result process grows linearly with the number of processes. In all combinations, with up to 7 processes there are less than 100 states in the result process. 
%%%% AR: Note that the split version also has only one merge?
%
%The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes. This is common with other systems based on inlining and/or template meta-programming, and is not prohibitive.
%
%On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes. The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream. The output of each merge process is the input of the next, forming a chain of merges. In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.
%
%When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product. During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition. The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result. 
%
%In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each. If we assume the two processes have about the same number of states then this is O($n^2$). Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused. 
%
%In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction. During fusion, the states for the consuming process are inlined at each occurrence of @push@. These states are legitimately different because at each occurence of @push@ the input channels of the merge process are in different channel states, and these channel states are included in the overall process state.

% -----------------------------------------------------------------------------
\subsection{Optimisation and Drop Instructions}
\label{s:Optimisation}
After we have fused two processes together, it may be possible to simplify the result before fusing in a third. Consider the result of fusing @group@ and @merge@ which we saw back in Fig.~\ref{fig:Process:Fused}. At labels @F1@ and @F2@ are two consecutive @jump@ instructions. The update expressions attached to these instructions are also non-interfering, which means we can safely combine these instructions into a single @jump@. In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code. The (PreferJump) clauses of Fig.~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.

Other @jump@ instructions like the one at @F5@ have no associated update expressions, and thus can be eliminated completely. Another simple optimization is to perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.

Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has fewer states. Provided we do not change the order of instructions that require synchronization with other processes (@pull@, @push@ or @drop@), the fusibility of the overall process network will not be affected.

Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable. In Fig.~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction. Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable. Values pulled from stream @sIn1@ are first stored in @b1@ before being copied to @v@ and @x1@. When the two processes to be fused share a common input stream, use of stream buffer variable allows one process to continue using the value that was last pulled from the stream, while the other moves onto the next one.

When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable. This is the case in Fig.~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@. In the result, the interleaved instructions from both source processes then share the same heap variable.

@@

% -----------------------------------------------------------------------------
% \begin{figure}
% \begin{minipage}[t]{0.10\textwidth}
% \end{minipage}
% \begin{minipage}[t]{0.4\textwidth}
% \begin{tikzpicture}
% \begin{axis}[
% % Hide the label on the second graph
% 	ylabel=Process states,
% 	xlabel=Pipelined combinators,
%   ymin=0, ymax=100,
% 	enlargelimits=0.01,
% 	ybar interval=0.7,
%   width=7.5cm, height=4.5cm,
% 	legend pos=north west,
% ]
% \addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
% \addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
% \legend{With merge, Without merge};
% \end{axis}
% \end{tikzpicture}
% \end{minipage}
% \quad
% \quad
% \quad
% \begin{minipage}[t]{0.4\textwidth}
% \begin{tikzpicture}
% \begin{axis}[
% %       ylabel=Process states,
%         xlabel=Parallel combinators,
%   ymin=0, ymax=100,
%         enlargelimits=0.01,
%         ybar interval=0.7,
%   width=7.5cm, height=4.5cm,
%         legend pos=north west,
% ]
% \addplot coordinates {(1,9) (2,27) (3,40) (4,53) (5,66) (6,79) (7,92)
%   % Last bar doesn't show for some reason, so need to add a dummy value for the next one
%     (8,1) };
% 
% \addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
% 
% \legend{With merge, Without merge};
% \end{axis}
% \end{tikzpicture}
% \end{minipage}
% 
% \caption{Maximum output process size for fusing all combinations of up to $n$ combinators.}
% \label{fig:bench:outputsize}
% \end{figure}
% 
% 
% % -----------------------------------------------------------------------------
% \begin{figure}
% 
% \begin{tikzpicture}
% \begin{axis}[
% 	ylabel=Process states,
% 	xlabel=Number of merges,
% %  ymode=log,
%   ymin=0, ymax=1500,
% 	enlargelimits=0.01,
% 	ybar interval=0.7,
%   width=7.5cm, height=4.5cm,
% 	legend pos=north west,
% ]
% % These are the values for splitting.
% % They are smaller than the 'chaining', but look much nicer on the linear graph.
% \addplot coordinates {(1,9) (2,42) (3,97) (4,196) (5,383) (6,746) (7,1461)
%   % (8,2880)
%   (8,1)
%   };
% 
% % These are the values for chaining
% % \addplot coordinates {(1,4) (2,48) (3,194) (4,760) (5,2814) (6,10064) (7,1) };
% 
% \end{axis}
% \end{tikzpicture}
% 
% \caption{Exponential blowup occurs when splitting or chaining merges together.}
% \label{fig:bench:exponential}
% \end{figure}

### 2017-05-04 2017mergingmerges/section/XX-FutureWork.tex
notes
typesetting to cut down
serious cutting

% We now discuss some of the shortcomings of the system, and extensions and future work to ameliorate this.

@@

The fusion algorithm treats all @case@ conditions as fully abstract, by exploring all possible combinations for both processes.
This can cause issues for processes that dynamically require only a bounded buffer, but the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.

For example, suppose two processes have the same case condition (@x > 0@).
The fusion algorithm will generate all four possibilities, including the contradictory ones such as where the first process has (@x > 0 = true@) and the second has (@x > 0 = false@).
If any require an unbounded buffer, the fusion algorithm will fail.

A possible extension is to somehow cull these contradictory states so that if it is statically known that a state is unreachable, it does not matter if it requires unbounded buffers.
It may be possible to achieve this with relatively little change to the fusion algorithm itself, by having it emit some kind of failure instruction rather than failing to produce a process.
A separate postprocessing step could then perform analyses and remove statically unreachable process states.
After postprocessing, if any failure instructions are reachable, fusion fails as before.

@@

The processes we have seen so far deal with infinite streams, but in practice most streams are finite. Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output. We have focussed on the infinite stream version because it is simpler to explain and prove, but the extensions required to support finite streams do not require substantial conceptual changes.

@@

For a combinator such as @take@ which only needs a finite prefix of the input stream, it may be useful to add an instruction that allows a pulling process to explicitly disconnect from an input channel. After disconnecting, the process no longer needs to be coordinated with the producer, leaving the producer free to push as often as the other consumers allow. We have implemented an early prototype that supports finite streams, our mechanized formalization does not cover it.

It is also possible to encode finite streams as infinite streams with an explicit end-of-stream marker (EOF) and case statements. However, this requires the fusion transform to analyse and reason about case statements and their predicates. It seems obvious that if two consumers read the same value and check if it is EOF, they are both going to give the same result, but this is of course undecidable in general. By making the structure of finite streams explicit and constraining processes to use finite streams in particular ways, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.

>>>>

%%% AR: Can we fit a few sentences here just to conclude, say we've described a fusion algorithm, with promising future work too

@@

The fusion algorithm treats all @case@ conditions as fully abstract by exploring all possible combinations for both processes.
This can cause issues for processes that dynamically require only a bounded buffer, but the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.

For example, if two processes have the same case condition (@x > 0@), the fusion algorithm generates all four possibilities, including contradictory ones where one process is true and the other is false.
If any require an unbounded buffer, the fusion algorithm fails.

One solution may be to cull contradictory states so that if it is statically known that a state is unreachable, it does not matter if it requires unbounded buffers.
It may be possible to achieve this with relatively little change to the fusion algorithm itself, by having it emit a failure instruction rather than failing to produce a process.
A separate postprocessing step could then remove statically unreachable process states.
After postprocessing, if any failure instructions are reachable, fusion fails as before.

@@

The processes we have seen so far deal with infinite streams, but in practice most streams are finite. Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output. We have focussed on the infinite stream version as it is simpler to explain and prove, but supporting finite streams does not require substantial conceptual changes.

@@

% For a combinator such as @take@ which only needs a finite prefix of the input stream, it may be useful to add an instruction that allows a pulling process to explicitly disconnect from an input channel. After disconnecting, the process no longer needs to be coordinated with the producer, leaving the producer free to push as often as the other consumers allow. Our implementation supports finite streams, but our mechanized formalization does not cover it.

It is also possible to encode finite streams as infinite streams with an explicit end-of-stream marker (EOF) and case statements. However, this requires the fusion transform to analyse and reason about case statements and their predicates. It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general. By making the structure of finite streams explicit and constraining how processes use finite streams, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.

### 2017-05-04 2017mergingmerges/section/XX-Proof.tex
attempt to fit in sigconf 12p. format
serious cutting

Our fusion system is formalized in Coq, and we have proved soundness of \ti{fusePair}: if the fused result process produces a particular sequence of values on its output channels then the two source processes may also produce that same sequence. Note that due to non-determinism of process execution the converse is not true in practice: just because the two concurrent processes can produce a particular output sequence does not mean the fused result process will as well --- the fused result process uses only one of the many possible orders.

@@

To aid mechanization, the Coq formalization has some small differences from the system presented in this paper. Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction. Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition. Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Fig.\ref{fig:Process:Eval:Feed}. Instead, we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process. This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.

>>>>

Our fusion system is formalized in Coq, and we have proved soundness of \ti{fusePair}: if the fused result process produces a particular sequence of values on its output channels then the two source processes may also produce that same sequence. The converse is not true, however: concurrent evaluation represents all possible interleavings, of which the fused process is only one.

@@

To aid mechanization, the Coq formalization has some small differences from the system presented in this paper. Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction. Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition. Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Fig.~\ref{fig:Process:Eval:Feed}. Instead, we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process. This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.
